Appendix

### 2024-10-16 manuscript_rmarkdown/manuscript_appendix.Rmd
Changes to appendix figure 1

```{=latex}
\begin{equation}
\phi_{\text{BC}}^{\lambda, x_0, s} (x_i) = 
\begin{cases}
\left( \left(\frac{x_i - x_0}{s} \right)^\lambda - 1 \right) / \lambda & \text{if } \lambda \neq 0\\
\log\left[\frac{x_i - x_0}{s}\right] & \text{if } \lambda = 0
\end{cases}
\end{equation}
```
Where $x_i - x_0 > 0$.
The location and scale-invariant Yeo-Johnson transformation is:

@@

We introduced per-sample weights $w_i$ to facilitate robust power transformations. The weighted log-likelihood of the location- and scale-invariant Box-Cox transformation is:
```{=latex}
\begin{equation}
\begin{split}
\mathcal{l}_{\text{rBC}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \lambda \left( \sum_{i=1}^n w_i \right) \log s + \left( \lambda - 1 \right) \sum_{i=1}^n w_i \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}
```
where $\mu_w$ and $\sigma^2_w$ are the weighted mean and weighted variance of the Box-Cox transformed feature $\phi_{\text{BC}}^{\lambda, x_0, s} (\mathbf{X})$, respectively:

@@

The first class of criteria is based on the empirical distribution function (EDF).
Transformation parameters are then fit through minimisation of the distance between the empirical distribution function $F_{\epsilon}$ and the cumulative density function (CDF) of the normal distribution $F_{\mathcal{N}}$.
Let $F_{\epsilon}\left(x_i \right) = \frac{i - 1/3}{n + 1/3}$ be the empirical probability of instance $i$.
The normal distribution is parametrised by location parameter $\mu$ and scale parameter $\sigma$, both of which have to be estimated from the data. 
For non-robust power transformation, $\mu$ and $\sigma$ are sample mean and sample standard deviation, respectively.
For robust power transformations, we estimate $\mu$ and $\sigma$ as Huber M-estimates of location and scale of the transformed feature $\phi^{\lambda, x_0, s} (\mathbf{X})$ [@Huber1981-su].

@@

The location, or mean, is weighted using weights $w_i$.
For non-robust transformations, $w_i = 1$.
Then, $sigma^2$ is defined as:
```{=latex}
\begin{equation}
\sigma^2 = \frac{\sum_{i=1}^n w_i \left(\phi^{\lambda, x_0, s} \left( x_i \right) - \mu \right)^2}{\sum_{i=1}^n w_i}
\end{equation}
```

@@

For the kurtosis test statistic we first compute [@DAgostino1990-kp, @Anscombe1983-nz]:

@@

Invariance of location- and scale-invariant power transformations was assessed using the optimisation criteria in (Appendix B)[Appendix B: Optimisation of transformation parameters].
This follows the simulation in the main manuscript, where MLE was used for optimization.
In short, we first randomly drew $10000$ values from a normal distribution: $\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{N}\left(0, 1\right)$, or equivalently $\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.5, 2\right)$.
The second distribution was a right-skewed normal distribution $\mathbf{X}_{\text{right}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.2, 2\right)$.
The third distribution was a left-skewed normal distribution $\mathbf{X}_{\text{left}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.8, 2\right)$.
We then computed transformation parameter $\lambda$ using the original definitions (equations \ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}) and the location- and scale-invariant definitions (equations \ref{eqn:box-cox-invariant} and \ref{eqn:yeo-johnson-invariant}) for each distribution using different optimisation criteria.
To assess location invariance, a positive value $d_{\text{shift}}$ was added to each distribution with $d_{\text{shift}} \in [1, 10^6]$.
Similarly, to assess scale invariance, each distribution was multiplied by a positive value $d_{\text{scale}}$, where $d_{\text{scale}} \in [1, 10^6]$.

@@

Applying conventional and invariant Box-Cox transformations to the latitude of houses in the Ames housing dataset [@De-Cock2011-jf] yielded the following results:
no transformation (residual sum $\sum r_i = 328$);
conventional transformation ($\lambda = 62.1$, $\sum r_i = 319$, $\mu_{BC} = 1.1 \cdot 10^{99}$, $\sigma_{BC} = 0.0 \cdot 10^{99}$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 96.0$, $\sum r_i = 319$, $\mu_{BC} = 6.2 \cdot 10^{153}$, $\sigma_{BC} = 0.3 \cdot 10^{153}$);
location- and scale-invariant transformation ($\lambda = 1.9$, $\sum r_i = 312$, $\mu_{BC} = 2.3$, $\sigma_{BC} = 0.9$);
and robust location- and scale-invariant transformation ($\lambda = 1.2$, $\sum r_i = 316$, $\mu_{BC} = 5.5$, $\sigma_{BC} = 1.4$.

@@

Results for Box-Cox transformations of features with outliers are shown in Figure \ref{fig:results-outlier-robustness-appendix}.

@@

```{r empirical-central-normality-test-appendix, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Type 1 error rate as function of the test statistic $\\tau_{\\text{ecn}}$ ",
  "for five datasets, with central portion $\\kappa=0.80$. ",
  "The type 1 error rate is computed from $m_d=10000$ randomly sampled features, ",
  "These are sampled from asymmetric generalized normal distributions, with and without outliers (Box-Cox and Yeo-Johnson), ",
  "or normal distributions with outliers (none). ",
  "The test statistic is computed as the average residual of each feature after (Box-Cox and Yeo-Johnson) robust location- and shift-invariant power transformation, or before (none)."
)

@@

The results are shown in Figure \ref{fig:empirical-central-normality-test-appendix} and Table \ref {tab:empirical-central-normality-appendix}.
These indicate that the test behaves similarly for the different datasets.
For low type 1 error rates the test statistic proposed in the main manuscript is more conservative than alternatives 
based on residuals after Box-Cox transformations of asymmetric generalised normally distributed features or on residuals from strictly normally distributed features.

>>>>

```{=latex}
\begin{equation}
\phi_{\text{BC}}^{\lambda, x_0, s} (x_i) = 
\begin{cases}
\left( \left(\frac{x_i - x_0}{s} \right)^\lambda - 1 \right) / \lambda & \text{if } \lambda \neq 0\\
\log\left[\frac{x_i - x_0}{s}\right] & \text{if } \lambda = 0
\end{cases}
\end{equation}
```
where $x_i - x_0 > 0$.
The location and scale-invariant Yeo-Johnson transformation is:

@@

Similarly to @Raymaekers2024-zf, sample weights $w_i$ are introduced to facilitate robust power transformations.
The weighted log-likelihood of the location- and scale-invariant Box-Cox transformation is:
```{=latex}
\begin{equation}
\begin{split}
\mathcal{l}_{\text{rBC}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \lambda \left( \sum_{i=1}^n w_i \right) \log s + \left( \lambda - 1 \right) \sum_{i=1}^n w_i \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}
```
where $\mu_w$ and $\sigma^2_w$ are the weighted mean and weighted variance of the Box-Cox transformed feature $\phi_{\text{BC}}^{\lambda, x_0, s} (\mathbf{X})$, respectively:

@@

The first class of criteria is based on the empirical distribution function (EDF).
Transformation parameters are then fit through minimisation of the distance between the empirical distribution function $F_{\epsilon}$ and the cumulative density function (CDF) of the normal distribution $F_{\mathcal{N}}$.
Let $F_{\epsilon}\left(x_i \right) = \frac{i - 1/3}{n + 1/3}$ be the empirical probability of instance $i$.
The normal distribution is parametrised by location parameter $\mu$ and scale parameter $\sigma$, both of which have to be estimated from the data. 
For non-robust power transformations, $\mu$ and $\sigma$ are sample mean and sample standard deviation, respectively.
For robust power transformations, we estimate $\mu$ and $\sigma$ as Huber M-estimates of location and scale of the transformed feature $\phi^{\lambda, x_0, s} (\mathbf{X})$ [@Huber1981-su].

@@

The location, or mean, is weighted using weights $w_i$.
For non-robust transformations, $w_i = 1$.
Then, $\sigma^2$ is defined as:
```{=latex}
\begin{equation}
\sigma^2 = \frac{\sum_{i=1}^n w_i \left(\phi^{\lambda, x_0, s} \left( x_i \right) - \mu \right)^2}{\sum_{i=1}^n w_i}
\end{equation}
```

@@

For the kurtosis test statistic we first compute [@DAgostino1990-kp; @Anscombe1983-nz]:

@@

Invariance of location- and scale-invariant power transformations was assessed using the optimisation criteria in [Appendix B](Appendix B: Optimisation of transformation parameters).
This follows the simulation in the main manuscript, where MLE was used for optimization.
In short, we first randomly drew $10000$ values from a normal distribution: $\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{N}\left(0, 1\right)$, or equivalently $\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.5, 2\right)$.
The second distribution was a right-skewed normal distribution $\mathbf{X}_{\text{right}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.2, 2\right)$.
The third distribution was a left-skewed normal distribution $\mathbf{X}_{\text{left}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.8, 2\right)$.
We then computed transformation parameter $\lambda$ using the original definitions (equations \ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}) and the location- and scale-invariant definitions (equations \ref{eqn:box-cox-invariant} and \ref{eqn:yeo-johnson-invariant}) for each distribution using different optimisation criteria.
To assess location invariance, a positive value $d_{\text{shift}}$ was added to each distribution with $d_{\text{shift}} \in [1, 10^6]$.
Similarly, to assess scale invariance, each distribution was multiplied by a positive value $d_{\text{scale}}$, where $d_{\text{scale}} \in [1, 10^6]$.

@@

Applying conventional and invariant Box-Cox transformations to the latitude of houses in the Ames housing dataset [@De-Cock2011-jf] yielded the following results:
no transformation (residual sum $\sum r_i = 328$);
conventional transformation ($\lambda = 62.1$, $\sum r_i = 319$, $\mu_{BC} = 1.1 \cdot 10^{99}$, $\sigma_{BC} = 0.0 \cdot 10^{99}$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 96.0$, $\sum r_i = 319$, $\mu_{BC} = 6.2 \cdot 10^{153}$, $\sigma_{BC} = 0.3 \cdot 10^{153}$);
location- and scale-invariant transformation ($\lambda = 1.9$, $\sum r_i = 312$, $\mu_{BC} = 2.3$, $\sigma_{BC} = 0.9$);
and robust location- and scale-invariant transformation ($\lambda = 1.2$, $\sum r_i = 316$, $\mu_{BC} = 5.5$, $\sigma_{BC} = 1.4$).

@@

Results for Box-Cox transformations of features with outliers are shown in Figure \ref{fig:experimental-results-outlier-robustness-appendix}.

@@

```{r empirical-central-normality-test-appendix, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Type I error rate as function of the test statistic $\\tau_{\\text{ecn}}$ ",
  "for five datasets, with central portion $\\kappa=0.80$. ",
  "The type I error rate is computed from $m_d=10000$ randomly sampled features, ",
  "These are sampled from asymmetric generalized normal distributions, with and without outliers (Box-Cox and Yeo-Johnson), ",
  "or normal distributions with outliers (none). ",
  "The test statistic is computed as the average residual of each feature after (Box-Cox and Yeo-Johnson) robust location- and shift-invariant power transformation, or before (none)."
)

@@

The results are shown in Figure \ref{fig:empirical-central-normality-test-appendix} and Table \ref {tab:empirical-central-normality-appendix}.
These indicate that the test behaves similarly for the different datasets.
For low type I error rates, the test statistic proposed in the main manuscript is more conservative than alternatives 
based on residuals after Box-Cox transformations of asymmetric generalised normally distributed features or on residuals from strictly normally distributed features.
